The Units of Permissionless Consensus: \\ Towards Mobile and Edge Computing

### 2022-10-12 theoretical-consensus/article/article.tex
add page 03

\section{Introduction}
% The very first letter is a 2 line initial drop letter followed
% by the rest of the first word in caps.
% 
% form to use if the first word consists of a single letter:
% \IEEEPARstart{A}{demo} file is ....
% 
% form to use if you need the single drop letter followed by
% normal text (unknown if ever used by the IEEE):
% \IEEEPARstart{A}{}demo file is ....
% 
% Some journals put the first two words in caps:
% \IEEEPARstart{T}{his demo} file is ....
% 
% Here we have the typical use of a "T" for an initial drop letter
% and "HIS" in caps to complete the first word.
\IEEEPARstart{L}{ong} has been the time when consensus started 
to be defined as a fundamental problem of distributed systems \cite{pease1980reaching, lamport2019byzantine, lamport1983weak}. 
Generally, consensus means reaching an agreement between multiple parties 
in the potential presence of faulty individuals. As per multiagent systems, 
interacting over computer networks, consensus is thought to be the result 
of a coordination effort such that those parties agree on some value at a 
given moment. Achieving consensus implies that the system shall be reliable 
and fault-tolerant. However, the consensus problem has been limited by some 
assumptions on the networks. The well-known "secure Byzantine-Fault-Tolerant 
multiparty consensus systems" that have been designed over the years are usually 
meant to work only with a set of known participants, faulty or not \cite{castro1999practical}. The 
other side of the coin is the permissionless consensus challenge, consisting of 
achieving agreement in an environment where the participants are unknown and 
untrusted \cite{nakamoto2008bitcoin, buterin2014next}. Plus, there are other intrinsic particularities of 
this type of networks, for example, their openness, and the lack of any kind of central 
authority. This adds another layer of complexity to the problem, as the participants 
are not only unknown and untrusted but can also join or leave the network at any time, 
freely choosing if they want to participate in the consensus protocol or not. Nevertheless, 
the problem of permissionless consensus can still be seen as a special case of the more 
general consensus and can still be formalized in the same way. In this paper, we will 
focus on consensus in permissionless systems, especially in the context of blockchain 
networks. We will reason about its meaningfulness, ultimately by trying to identify the units that 
may underpin consensus. We will also discuss the current state-of-the-art, with a 
particular interest in the shift of the consensus layer of these distributed networks 
to mobile and edge computing environments, for which computationally expensive consensus 
algorithms are impractical and unfair.

@@

The establishment of a definition for the problem of reaching agreement in a distributed 
system was pioneered by Lamport et al. in \cite{pease1980reaching}. 
The authors defined consensus as the problem of agreeing on a single value among a set of processes, 
in the presence of faulty entities. The first consensus algorithms were designed for 
synchronous systems, where the communication between the processes is reliable, and the 
delay is bounded. However, these initial attempts failed to cover the different types of 
faulty behaviour. Along with the establishment of the famous Byzantine Generals Problem, 
the first solutions, not only for dealing with treason, but also for unreliable communication 
channels, or any other kind of arbitrary Byzantine behaviour, were 
also proposed by Lamport in \cite{lamport2019byzantine,lamport1983weak}. The solution was a synchronous 
mechanism that used a set of leaders to reach consensus. Multiple practical implementations 
and optimizations to this solution have been proposed in the literature.

@@

Technically, permissionless environments allow for larger networks
that depict lower connectivity between the participants. Operationally,
everything is expected to happen in an asynchronous or partially synchronous fashion, 
and the number of transactions is expected to be smaller than the 
permissioned counterparts. Nevertheless, participation is free, and the
governance is not centralized, but rather distributed and public. 
The identity of the participants is secured or semi-secure as it often relies on
pseudonymity for protecting the nodes' identity, while enabling full transparency 
in regards to the rest of the network content and operation \cite{xiao2019distributed}.
The goal of permissionless consensus, as generally for any consensus protocol, is to reach
agreement on a single value, or a set of values. However, due to the nature of the
protocols, the values that are agreed upon end up establishing the serialization of the
transactions that are executed in the network, and so establishing time consciousness and 
total order of the events.

@@

Also described in \cite{survey-dist-consensus}, very concisely, the 
way to achieve an operating protocol, as seen in the mainstream blockchain
networks, is by chaining the generation of the agreeable value, 
in this particular case, a block and its proof, with the dissemination of 
the information to the rest of the network, followed by the eventual validation and 
acceptance of the block by the rest of the nodes. This is the moment when consensus is reached.
Nevertheless, during the whole process, a fair and somewhat predictable incentive mechanism 
is also needed, that rewards the participants for their honest effort in reaching consensus, and
punishes the ones that are not behaving correctly. These incentives are of major importance
in this very context of permissionless consensus. 
These building blocks form the basis of the inner functioning of Bitcoin itself \cite{nakamoto2008bitcoin},
and are replicated with some variations in the other permissionless networks \cite{buterin2014next}.
For example, there are some networks that do not have a proof-of-work mechanism,
but rather a proof-of-stake, or a proof-of-space, or a proof-of-burn mechanism, etc.
when it comes to the generation of the block and its existential and later verifiable proof.
All of the other pillars are generally the same \cite{survey-dist-consensus}.
The next sections will try to give a more detailed overview of these 
block generation proof units.

@@

In the classical Nakamoto consensus protocol, the generation of a block, to be proposed
for further network agreement, complies with the unit of computational work needed to
create, or rather find, a verifiable proof of the effort spent on creating the block \cite{nakamoto2008bitcoin}. 
This essentially requires brute forcing the search for a cryptographic hash value for the
aggregation of the block information with a nonce, such that this hash value satisfies
a difficulty threshold (see Algorithm \ref{alg:BlockGeneration}), which gets adjusted dynamically over time, to maintain the network overall 
requirement for the block generation interval \cite{survey-dist-consensus}.

There are a couple of particularities that need to be mentioned. First, the block generation
is of probablistic nature, as the nonce is a random value, and the hash function is a one-way function.
Allied to this, more hashing power is not easily obtained, which allows for tackling Sybil attacks,
despite the permissionless and pseudonymity nature of the network. Plus, the incentive mechanism
also plays a role in this regard, by naturally incentivizing honest participation.
The block generation interval, forged by the adjustment of the difficulty
threshold, is also of major importance because it reduces the fork probability and
allows for the timely propagation of proposed blocks, guaranteeing consensus
in the presence of an honest majority of the hashing power \cite{garay2015bitcoin}.
This probabilitic finality shifts the 1/3 BFT threshold to a 1/2 threshold.

@@

\subsection{Trade-offs and trillema}

Despite the harmony of all the aforementioned clever characteristics,
the proof-of-work mechanism has some well known drawbacks. 
First, the block generation is computationally expensive, and so the
energy consumption associated with it is also high. This is caused by the
dynamically adjusted difficulty threshold, which is a function of the block 
generation interval. To maintain the 10 minutes interval, the hashing difficulty
increases, and so too the hashing power and the energy needed for it. 
However, reducing the block generation interval would increase the
probability of forks, which results in a more frequent waste of useful work,
loosening the security of the network. Increasing the block size would also
have the same security effect due to the increased propagation delay.

As pointed in the literature, the design of these consensus mechanisms
shall aim for a protocular choice between a set of properties that form a trillema (see Fig. \ref{fig:trillema}): 
Security, Scalability, and Decentralization \cite{survey-dist-consensus}.
Summarizing, ...

\begin{figure}[h]
  \centering
  \includegraphics[width=\columnwidth]{trillema}
  \caption{From \cite{survey-dist-consensus}, the blockchain consensus trillema.}
  \label{fig:trillema}
\end{figure}

>>>>

\section{Introduction}
% The very first letter is a 2 line initial drop letter followed
% by the rest of the first word in caps.
% 
% form to use if the first word consists of a single letter:
% \IEEEPARstart{A}{demo} file is ....
% 
% form to use if you need the single drop letter followed by
% normal text (unknown if ever used by the IEEE):
% \IEEEPARstart{A}{}demo file is ....
% 
% Some journals put the first two words in caps:
% \IEEEPARstart{T}{his demo} file is ....
% 
% Here we have the typical use of a "T" for an initial drop letter
% and "HIS" in caps to complete the first word.
\IEEEPARstart{L}{ong} has been the time when consensus started 
to be defined as a fundamental problem of distributed systems \cite{pease1980reaching, lamport2019byzantine, lamport1983weak}. 
Generally, consensus means reaching an agreement between multiple parties 
in the potential presence of faulty individuals. As per multi-agent systems, 
interacting over computer networks, consensus is thought to be the result 
of a coordination effort such that those parties agree on some value at a 
given moment. Achieving consensus implies that the system shall be reliable 
and fault-tolerant. However, the consensus problem has been limited by some 
assumptions on the networks. The well-known "secure Byzantine-Fault-Tolerant 
multiparty consensus systems" that have been designed over the years are usually 
meant to work only with a set of known participants, faulty or not \cite{castro1999practical}. The 
other side of the coin is the permissionless consensus challenge, consisting of 
achieving agreement in an environment where the participants are unknown and 
untrusted \cite{nakamoto2008bitcoin, buterin2014next}. Plus, there are other intrinsic particularities of 
this type of networks, for example, their openness, and the lack of any kind of central 
authority. This adds another layer of complexity to the problem, as the participants 
are not only unknown and untrusted but can also join or leave the network at any time, 
freely choosing if they want to participate in the consensus protocol or not. Nevertheless, 
the problem of permissionless consensus can still be seen as a special case of the more 
general consensus and can still be formalized in the same way. In this paper, we will 
focus on consensus in permissionless systems, especially in the context of blockchain 
networks. We will reason about its meaningfulness, ultimately by trying to identify the units that 
may underpin consensus. We will also discuss the current state-of-the-art, with a 
particular interest in the shift of the consensus layer of these distributed networks 
to mobile and edge computing environments, for which computationally expensive consensus 
algorithms are impractical and unfair.

@@

The establishment of a definition for the problem of reaching agreement in a distributed 
system was pioneered by Lamport et al. in \cite{pease1980reaching}. 
The authors defined consensus as the problem of agreeing on a single value among a set of processes, 
in the presence of faulty entities. The first consensus algorithms were designed for 
synchronous systems, where the communication between the processes is reliable, and the 
delay is bounded. However, these initial attempts failed to cover the different types of 
faulty behavior. Along with the establishment of the famous Byzantine Generals Problem, 
the first solutions, not only for dealing with treason, but also for unreliable communication 
channels, or any other kind of arbitrary Byzantine behavior, were 
also proposed by Lamport in \cite{lamport2019byzantine,lamport1983weak}. The solution was a synchronous 
mechanism that used a set of leaders to reach consensus. Multiple practical implementations 
and optimizations to this solution have been proposed in the literature.

@@

Technically, permissionless environments allow for larger networks
that depict lower connectivity between the participants. Operationally,
everything is expected to happen in an asynchronous or partially synchronous fashion, 
and the number of transactions is expected to be smaller than the 
permissioned counterparts. Nevertheless, participation is free, and the
governance is not centralized, but rather distributed and public. 
The identity of the participants is secured or semi-secure as it often relies on
pseudonymity for protecting the nodes' identity, while enabling full transparency 
in regard to the rest of the network's content and operation \cite{xiao2019distributed}.
The goal of permissionless consensus, as generally for any consensus protocol, is to reach
agreement on a single value, or a set of values. However, due to the nature of the
protocols, the values that are agreed upon end up establishing the serialization of the
transactions that are executed in the network, and so establishing time consciousness and 
total order of the events.

@@

Also described in \cite{survey-dist-consensus}, very concisely, the 
way to achieve an operating protocol, as seen in the mainstream blockchain
networks, is by chaining the generation of the agreeable value, 
in this particular case, a block and its proof, with the dissemination of 
the information to the rest of the network, followed by the eventual validation and 
acceptance of the block by the rest of the nodes. This is the moment when consensus is reached.
Nevertheless, during the whole process, a fair and somewhat predictable incentive mechanism 
is also needed, that rewards the participants for their honest effort in reaching consensus, and
punishes the ones that are not behaving correctly. These incentives are of major importance
in this very context of permissionless consensus. 
These building blocks form the basis of the inner functioning of Bitcoin itself \cite{nakamoto2008bitcoin},
and are replicated with some variations in the other permissionless networks \cite{buterin2014next}.
For example, there are some networks that do not have a proof-of-work mechanism,
but rather a proof-of-stake, or a proof-of-space, or a proof-of-burn mechanism, etc.
when it comes to the generation of the block and its existential and later verifiable proof.
All the other pillars are generally the same \cite{survey-dist-consensus}.
The next sections will try to give a more detailed overview of these 
block generation proof units.

@@

@@

In the classical Nakamoto consensus protocol, the generation of a block, to be proposed
for further network agreement, complies with the unit of computational work needed to
create, or rather find, a verifiable proof of the effort spent on assembling the block \cite{nakamoto2008bitcoin}. 
This essentially requires brute forcing the search for a cryptographic hash value for the
aggregation of the block information with a nonce, such that this hash value satisfies
a difficulty threshold (see Algorithm \ref{alg:BlockGeneration}), which gets adjusted dynamically over time, to maintain the network overall 
requirement for the block generation interval \cite{survey-dist-consensus}.

There are a couple of particularities that need to be mentioned. First, the block generation
is of probabilistic nature, as the nonce is a random value, and the hash function is a one-way function.
Allied to this, more hashing power is not easily obtained, which allows for tackling Sybil attacks,
despite the permissionless and pseudonymity nature of the network. Plus, the incentive mechanism
also plays a role in this regard, by naturally incentivizing honest participation.
The block generation interval, forged by the adjustment of the difficulty
threshold, is also of major importance because it reduces the fork probability and
allows for the timely propagation of proposed blocks, guaranteeing consensus
in the presence of an honest majority of the hashing power \cite{garay2015bitcoin}.
This probabilistic finality shifts the 1/3 BFT threshold to a 1/2 threshold.

\subsection{Trade-offs and trilemma}

Despite the harmony of all the aforementioned clever characteristics,
the proof-of-work mechanism has some well known drawbacks. 
First, the block generation is computationally expensive, and so the
energy consumption associated with it is also high. This is caused by the
dynamically adjusted difficulty threshold, which is a function of the block 
generation interval. For example, in the Bitcoin network, to maintain the 10 
minutes interval, the hashing difficulty
increases with the increase of the hashing power, and so too the energy needed for it. 
However, reducing the block generation interval would increase the
probability of forks, which results in a more frequent waste of useful work,
loosening the security of the network. Increasing the block size would also
have the same security effect due to the increased propagation delay.

As pointed in the literature, the design of these consensus mechanisms
shall aim for a protocolar choice between a set of properties that form a trilemma (see Fig. \ref{fig:trilemma}): 
Security, Scalability, and Decentralization.
Summarizing, relaxing the security requirements may allow for more scalability, both of which,
consequently, have hands tied with decentralization. These trade-offs are of practical
consideration when defining the network goals and use cases \cite{survey-dist-consensus}.

\begin{figure}[h]
  \centering
  \includegraphics[width=\columnwidth]{trilemma}
  \caption{From \cite{survey-dist-consensus}, the blockchain consensus trilemma and its inner relations.}
  \label{fig:trilemma}
\end{figure}
